---
title: "10-import"
title-block-banner: true
author:
  - name: Jinlong Ru
    orcid: 0000-0002-6757-6018
date: 2025-04-04
toc: true
toc-depth: 4
number-sections: true
code-fold: true
code-line-numbers: true
code-tools: true
format: 
  html:
    embed-resources: true
    smooth-scroll: true
    page-layout: full
reference-location: section
citation-location: document
params:
  name: "10-import"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

The purpose of this document is to import the metadata and results from ViroProfiler and other tools to a SQLite database for further analysis.

```{r setup}
#| message: false
#| include: false
#| warning: false
wd <- "analyses"
if (basename(getwd()) != wd) {
  setwd(here::here(wd))
}
params <- list(name = "10-import")
here::i_am(paste0(params$name, ".qmd"), uuid = "17a61b94-8453-4573-b2fd-e19f8ef3738f")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r packages}
#| message: false
suppressPackageStartupMessages({
  library(here)
  library(conflicted)
  library(tidyverse)
  library(data.table)
  devtools::load_all()
})

con <- load_db()
```

## Tasks

NOTE: 

1. vOTU IDs are affected by the `map-aniclst.tsv` file.
2. VC IDs are affected by the `vcontact3-final_assignments.csv` file.
3. The `study_id` is the same as the `sample_id`, except for the samples from the iMeta2024 paper.

Viral contigs should satisfy the following criteria:

1. `genomad_virus_score` >= 0.8 & `virsorter2_score` >= 0.9

Viral-host pairs should satisfy the following criteria:

1. iPHoP `Confidence.score` >= 90


```{r metadata}
df_metadata <- openxlsx::read.xlsx(here(path_raw, "pc028e3-metadata.xlsx"), sheet = "metadata") %>% 
  # dplyr::select(c("study_id", "project_id", "sample_host", "geo_loc_name", "sample_id")) %>%
  dplyr::distinct()
```

```{r import hpc results}
df_vibrant <- parse_vibrant_quality(here(path_data, "vibrant-genome_quality.tsv"))
df_genomad <- parse_genomad_summary(here(path_data, "genomad-summary.tsv")) %>% 
  dplyr::select(-c("realm", "kingdom", "phylum", "class", "order", "family", "genus", "species"))
df_taxonomy <- fread(path_source("09-taxonomy", "taxonomy_formatted.tsv"))
df_virsorter2 <- parse_virsorter2(here(path_data, "virsorter2-score.tsv"), here(path_data, "virsorter2-category.csv"))
df_checkv <- parse_checkv_quality(here(path_data, "checkv-quality_summary.tsv"))
df_ctg2vOTU <- parse_aniclst(here(path_data, "map-aniclst.tsv"))
df_bacphlip <- parse_bacphlip(here(path_data, "bacphlip.tsv.gz"))
df_vcontact3 <- parse_vcontact3(here(path_data, "vcontact3-final_assignments.csv.gz"))
df_vcontact3_qry <- df_vcontact3 %>%
  dplyr::filter(Reference == FALSE) %>%
  dplyr::select(-c("Reference")) %>%
  data.table::setnames("GenomeName", "contig_repid")


# The df_ctg2asb table is variable between different projects, so we need to parse it separately
df_ctg2asb <- fread(here(path_data, "map-contig2asb.tsv"), header = FALSE) %>% 
  data.table::setnames(c("contig_id", "assembly_id")) %>% 
  dplyr::mutate(sample_id = stringr::str_replace_all(assembly_id, "ma2024imeta__", "ma2024imeta_")) %>%
  dplyr::mutate(sample_id = stringr::str_replace_all(sample_id, "__.*", "")) %>%
  # some sample_id contains "_clean", remove it
  dplyr::mutate(sample_id = stringr::str_replace_all(sample_id, "_clean", ""))
```

```{r merge all tables}
df_metadata_host <- df_metadata %>% 
  dplyr::select(c("study_id", "sample_host")) %>% 
  dplyr::distinct()


df_contig_annotation <- df_ctg2vOTU %>%
  dplyr::left_join(df_ctg2asb, by = "contig_id") %>% 
  dplyr::left_join(df_checkv, by = "contig_id") %>% 
  dplyr::left_join(df_taxonomy, by = "contig_id") %>%
  dplyr::inner_join(df_genomad, by = "contig_id") %>% 
  dplyr::inner_join(df_virsorter2, by = "contig_id") %>%
  dplyr::left_join(df_vibrant, by = "contig_id") %>%
  dplyr::left_join(df_bacphlip, by = "contig_id") %>%
  dplyr::left_join(df_vcontact3_qry, by = "contig_repid") %>%
  # create new column "lifestyle", set value to "temperate" if "vibrant_type" is "lysogenic" or "genomad_topology" is "Provirus" or "checkv_provirus" is "Yes" or "bacphlip_lifestyle" is "temperate", otherwise set to "virulent"
  dplyr::mutate(lifestyle = dplyr::case_when(
    vibrant_type == "lysogenic" | genomad_topology == "Provirus" | checkv_provirus == "Yes" | bacphlip_lifestyle == "temperate" ~ "temperate",
    TRUE ~ "virulent")) %>%
  # sort by contig_id
  dplyr::arrange(contig_id) %>%
  # fill missing values in sample_id with previous non-NA value
  tidyr::fill(sample_id, .direction = "down") %>%
  dplyr::mutate(study_id = stringr::str_replace_all(sample_id, "_.*", "")) %>%
  dplyr::mutate(assembly_id = ifelse(is.na(assembly_id), str_c(sample_id, "__missing"), assembly_id)) %>%
  # put "contig_id" and "study_id" at the beginning of the data frame
  dplyr::select(c("contig_id", "study_id", everything())) %>%
  # only keep high confidence viral contigs
  dplyr::filter(genomad_virus_score >= 0.8 & virsorter2_score >= 0.9) %>% 
  # add bird host columns
  dplyr::left_join(df_metadata_host, by = "study_id") %>% 
  pivot_wider(names_from = sample_host, values_from = sample_host, values_fn = function(x) 1, values_fill = 0, names_prefix = "host_")
  

# Viral-host prediction by iPHoP
df_host <- fread(here(path_data, "iphop-prediction_to_genome.csv.gz")) %>%
  data.table::setnames(colnames(.), make.names(colnames(.))) %>%
  data.table::setnames("Virus", "contig_id") %>%
  dplyr::filter(Confidence.score >= 90) %>%
  dplyr::filter(contig_id %in% df_contig_annotation$contig_id)

# parse DRAM-v AMG results
df_amg_dramv <- read_amg(here(path_data, "dramv-amg.tsv"), "dramv")
df_amg_vibrant <- read_amg(here(path_data, "vibrant-amg.tsv"), "vibrant")
```

```{r empty assembly_id due to missing assembly results}
df_asb_empty <- df_contig_annotation %>%
  dplyr::filter(str_detect(assembly_id, "__missing"))

# save df_asb_empty to csv
fwrite(df_asb_empty, path_target("df_asb_empty.tsv"), sep = "\t")
```

```{r filter high-quality VCs for downstream analysis}
# filter by the selected column
df_contig_annotation_filtered <- df_contig_annotation %>% 
  # dplyr::filter(checkv_quality_score %in% c(1,2,3,4)) %>% 
  dplyr::filter(genomad_virus_score >= 0.9) %>% 
  dplyr::filter(checkv_contamination <= 5) %>% 
  dplyr::filter(virsorter2_score >= 0.9) %>% 
  dplyr::filter(vibrant_quality_score %in% c(1,2,3,4)) %>% 
  # sort by checkv_quality_score 
  dplyr::arrange(desc(checkv_completeness)) %>% 
  dplyr::filter(checkv_completeness >= 20)



df_contig_annotation_host <- df_contig_annotation %>% 
  dplyr::select(c("vOTU_id", starts_with("host_"))) %>% 
  # aggregate by vOTU_id, sum the values
  dplyr::group_by(vOTU_id) %>%
  dplyr::summarise(across(everything(), ~ sum(.x, na.rm = TRUE))) %>% 
  dplyr::ungroup()

# df_contig_annotation_host2 <- df_contig_annotation_host %>% 
  # dplyr::mutate(across(starts_with("host_"), ~ ifelse(. > 0, 1, 0)))

# vOTU representative contigs
df_vOTUrep_filtered <- df_contig_annotation_filtered %>% 
  # remove duplicates by vOTU_id, only keep the first one
  dplyr::distinct(vOTU_id, .keep_all = TRUE) %>% 
  # remove columns that contains "host_" in the name
  dplyr::select(-dplyr::matches("host_")) %>% 
  dplyr::left_join(df_contig_annotation_host, by = "vOTU_id") %>% 
  # vOTU prevalence in all hosts
  dplyr::rowwise() %>% 
  dplyr::mutate(vOTU_prevalence = sum(c_across(starts_with("host"))>0)) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(-c("contig_id", "study_id", "assembly_id", "sample_id"))


mean(df_vOTUrep_filtered$checkv_completeness, na.rm = TRUE)

df_amg_dramv_filtered <- dplyr::filter(df_amg_dramv, contig_id %in% df_contig_annotation_filtered$contig_id)
df_amg_vibrant_filtered <- dplyr::filter(df_amg_vibrant, contig_id %in% df_contig_annotation_filtered$contig_id)
df_host_filtered <- dplyr::filter(df_host, contig_id %in% df_contig_annotation_filtered$contig_id)
df_vcontact_filtered <-  dplyr::filter(df_vcontact3, vc_id %in% df_contig_annotation_filtered$vc_id)
df_metadata_filtered <-  dplyr::filter(df_metadata, study_id %in% df_contig_annotation_filtered$study_id)

# write filtered data to sqlite
DBI::dbWriteTable(con, "amg_dramv_filtered", df_amg_dramv_filtered, overwrite = TRUE)
DBI::dbWriteTable(con, "amg_vibrant_filtered", df_amg_vibrant_filtered, overwrite = TRUE)
DBI::dbWriteTable(con, "host_filtered", df_host_filtered, overwrite = TRUE)
DBI::dbWriteTable(con, "vcontact_filtered", df_vcontact_filtered, overwrite = TRUE)
DBI::dbWriteTable(con, "contig_annotation_filtered", df_contig_annotation_filtered, overwrite = TRUE)
DBI::dbWriteTable(con, "sample_metadata_filtered", df_metadata_filtered, overwrite = TRUE)
DBI::dbWriteTable(con, "contig_annotation_filtered_vOTU", df_vOTUrep_filtered, overwrite = TRUE)
```



```{r protein annotation}
df_eggnog <- fread(here(path_data, "anno_eggnog.tsv.gz")) %>% 
  dplyr::mutate(contig_id = str_replace_all(query, "_.*", "")) %>% 
  dplyr::rename(protein_id = query) %>%
  dplyr::select(c(contig_id, everything()))

df_dramv <- fread(here(path_data, "anno_dramv.tsv.gz")) %>% 
  dplyr::mutate(contig_id = str_replace_all(scaffold, "-cat_.*", "")) %>% 
  dplyr::mutate(protein_id = str_replace_all(V1, "-cat_.*_", "_")) %>%
  dplyr::select(c(contig_id, protein_id, everything()))

DBI::dbWriteTable(con, "anno_prot_eggnog", df_eggnog, overwrite = TRUE, row.names = FALSE)
DBI::dbWriteTable(con, "anno_prot_dramv", df_dramv, overwrite = TRUE, row.names = FALSE)
```


## Files written

These files have been written to the target directory, `r paste0("data/", params$name)`:

```{r list-files-target}
DBI::dbWriteTable(con, "vcontact", df_vcontact3, overwrite = TRUE)
DBI::dbWriteTable(con, "contig_annotation", df_contig_annotation, overwrite = TRUE)
DBI::dbWriteTable(con, "host", df_host, overwrite = TRUE)
DBI::dbWriteTable(con, "sample_metadata", df_metadata, overwrite = TRUE)
DBI::dbWriteTable(con, "amg_dramv", df_amg_dramv, overwrite = TRUE)
DBI::dbWriteTable(con, "amg_vibrant", df_amg_vibrant, overwrite = TRUE)
# DBI::dbExecute(con, "VACUUM")
DBI::dbDisconnect(con)
projthis::proj_dir_info(path_target(), tz = "CET")
```
